 Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings

### 2020-07-18 paper.Rmd
add marker gene example

First, we downloaded the raw mouse retinal single cell RNA-seq data from
@Macosko2015-ot using the `scRNAseq` Bioconductor package [@scRNAseq-d]. We
have followed a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed QC using the the `scater`
by removing cells with high proportion of mitochondrial gene expression and low
numbers of genes detected, we log-transformed and normalised the expression 
values and  finally selected highly variable genes (HVGs) using `scran` 
[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
used Louvain clustering to generate clusters [@Blondel2008-bx]. 
To check and verify the clustering we construct a liminal view. 
We tour the first five PCs (approximately 20% of the variance in expression),
along side the tSNE view which was computed from all 25 PCs. Due to latency
of the interface we do a ten percent weighted sample of rows based on
cluster membership.

- compare to just using the HVGs themselves with the tour
- neighbourhood based brushing to check how diffuse or distorted clusters are
- different distance metrics?

>>>>

First, we downloaded the raw mouse retinal single cell RNA-seq data from
@Macosko2015-ot using the `scRNAseq` Bioconductor package [@scRNAseq-d]. We
have followed a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed QC using the the `scater`
by removing cells with high proportion of mitochondrial gene expression and low
numbers of genes detected, we log-transformed and normalised the expression 
values and  finally selected highly variable genes (HVGs) using `scran` 
[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
used Louvain clustering to generate clusters [@Blondel2008-bx].

To check and verify the clustering we construct a `liminal` view. 
We tour the first five PCs (approximately 20% of the variance in expression),
along side the tSNE view which was computed from all 25 PCs. 
 Due to latency of the interface we do weighted sample of rows based on
cluster membership, leaving us with approximately 10 per cent of the original
data size - 4,590 cells. Although this is not ideal, it still allows us to
get a sense of the shape of the clusters as seen in the video below:

< embedded video mouse 1 >

In the following video we can observe that the embedding has mostly captured the
clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
linked brushing. As expected avoids tSNE mitigates the crowding problem that is
an issue for tour in this case, where points in clusters 2,4,6, and 11 are
clumped together in tour view, but are blown up in the embedding view. 
The tour appears to form a tetrahedron like shape, with points lying on
surface and along the vertices of the the tetrahedron in 5-$d$ PCA space -
a phenomena that has also been observed in @Korem2015-af. Brushing on
the tour view, reveals points in cluster 9 that are diffuse in the embedding 
view, points in cluster 9 are relatively far away and spread apart from other clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding.

Next, we identify marker genes for clusters using one sided Welch t-tests with
a minimum log fold change of one as recommended by @Amezquita2020-at, which
uses the testing framework from @McCarthy2009-qx. We select the top 10 marker genes that are upregulated in cluster 2, which was one of the clumped clusters
when we toured on principal components. Here the tour becomes an alternative
to a standard heatmap view for assessing shared markers; the basis
generation (shown as the biplot on the left view) reflects the relative 
weighting of each gene. We run the tour directly on the 
log normalised expression values using the same subset as before:

< embedded video mouse 2 >

We see that the expression of the marker genes using t, appear to separate
the previously clumped clusters 2,4,6, and 11 from the other clusters, 
indicating that these genes are expressed in all four clusters, and the 
magnitude of expression in these genes is similar accross these groups, 
consistent with the results of marker gene analysis.
